---
title: "Tidy indoctrination"
author: "Monica Thieu"
date: "July 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(tidyverse)
```

## Goals for this vignette

* Demonstrate (what I think are) core features of the tidyverse
* Illustrate psychology use cases for tidyverse functions
* Hopefully convince you to join the cult of tidy

We shall do this by walking through a day in the life of a tidyverse user (me, I guess). This vignette will be set up in a sort of infomercial style, with a BEFORE TIDY (the grayscale part where the hapless folks in the infomercial cannot do a dang thing) and AFTER TIDY section (the nice color part where the gadget makes their life so much better).

My goal here is twofold:

* Map strategies that you already implement in a non-tidy way to their tidy counterparts
* Demonstrate possible pain points in non-tidy methods that are resolved in the tidyverse

### Please keep in mind!

This vignette is **not** intended to be an exhaustive walkthrough of all functions from the indicated `tidyverse` packages.

For reference, please see the excellent reference pages online for the relevant package. For a more detailed self-teaching experience, please refer to the incredible [R for Data Science online textbook](http://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham, which teaches using `tidyverse` functions.

## Quick reminder of key functions from various pkgs

Remember that each of these pkgs contain other functions too! Often these other functions are more specific use cases of the primary functions that crop up often enough that you'd want a separate function for them.

To see all the functions in each pkg, visit their reference pages at tidyverse.org ! I love these reference pages so much.

* `%>%`
* `tidyr`
    + `gather()` & `spread()` for making long and wide
    + `separate()` & `unite()` for fixing key columns (pre/post-processing for gather/spread)
    + `expand()`, `crossing()`, `nesting()`-- like `expand.grid()` but a little more predictable
    + `nest()`/`unnest()` for tibble-ception
* `dplyr`
    + `select()`, `filter()` for subsetting and reordering dfs/tibbles
    + `group_by()` for... grouping
    + `mutate()` for your column manipulation workhorse
    + `summarize()` for squishing multiple rows down to summary statistics
    + `arrange()` to sort dfs/tibbles more easily
* `purrr`
    + all the `map()` functions: like `apply()` but tidy
    
## Brief intro to the pipe

Before we begin...

Enter the pipe `%>%`! The pipe does one simple, but key, thing: **takes the object on the left-hand side and feeds it into the first argument of the function on the right-hand side.** This means that:

* `a %>% foo()` is equivalent to `foo(a)`. Fine and good
* `a %>% foo() %>% bar(arg = TRUE)` is equivalent to `bar(foo(a), arg = TRUE)`. Now, nested function calls read left-to-right!
* Most common use case: `df_new <- df_old %>% foo() %>% bar(arg = TRUE) %>% baz()` is equivalent to `df_new <- baz(bar(foo(df_old), arg = TRUE))`. Now, you can chain a series of preprocessing commands to operate on a dataframe all at once, and easily read those commands as typed in your script. No more accidentally not running some key preprocessing command that causes later code to break!

Note that #pipelife requires functions to:

* take as their first argument the object to be operated upon
* return the same object (or an analog of said), but now operated upon

Essentially all functions from the tidyverse are pipe-safe, but bear this in mind when trying to incorporate functions from base R or other packages into your tidy new world.

## Initializing fake data

We'll use a simulated dataset for this vignette, so you don't need to worry about any dependencies involving datasets you don't have access to while you're following along.

```{r}
# 20 fake subjects, 50 fake trials per subject

# Will simulate the person-level variables FIRST,
# then expand to simulate the trial-level variables
raw <- tibble(id = 1L:20L,
              age = sample(18L:35L, size = 20, replace = TRUE),
              # assuming binary gender for the purposes of this simulation
              gender = sample(c("male", "female"), size = 20, replace = TRUE)) %>%
  # simulating some "questionnaire" scores; person-level
  mutate(q_1 = rnorm(n = n(), mean = 30, sd = 10),
              q_2 = rnorm(n = n(), mean = 30, sd = 10),
              q_3 = rnorm(n = n(), mean = 30, sd = 10)) %>%
  # slice() subsets rows by position; you can use it to repeat rows by repeating position indices
  slice(rep(1:n(), each = 50)) %>%
  # We'll get to this in a bit--this causes every "group"
  # aka every set of rows with the same value for "id", to behave as an independent df
  group_by(id) %>%
  # I just want to have a column for "trial order", I like those in my task data
  mutate(trial_num = 1:n(),
         # Each subject sees half OLD and half NEW trials in this recognition memory task
         is_old = rep(0L:1L, times = n()/2),
         # I'm shuffling the order of "old" and "new" trials in my fake memory task
         is_old = sample(is_old),
         # This will generate binary "old"/"new" responses corresponding roughly to a d' of 1
         # yep, everyone has the same d' today
         response = if_else(is_old == 1,
                            rbinom(n = n(), size = 1, prob = 0.7),
                            rbinom(n = n(), size = 1, prob = 0.3)),
         rt = rnorm(n = n(), mean = 3, sd = 1.5))
  
```


## Loading data

`readr` is great

### Non-tidy way

```{r}
wm_raw_untidy <- read.csv("~/Dropbox/MMT/data/MMT_filterObjs_all_final.csv", header = TRUE, stringsAsFactors = FALSE)

ltm_raw_untidy <- read.csv("~/Dropbox/MMT/data/MMT_testTargets_all_final.csv", header = TRUE, stringsAsFactors = FALSE)
```

### Tidy way

Don't forget to plug `read_csv_multi()` from this package

```{r, eval = FALSE}
data <- read_csv("/path/to/data.csv")
```

### Key differences

* in `readr`, header always `TRUE`, stringsAsFactors always `FALSE`
* `readr` always returns a tibble--good habits for later

### A taste of the possible

`googledrive` but it looks like you have to run through a temp mode of downloading the newest version from Google Drive and then reading that in from disk. OH but you could write a wrapper for this...

## Visually inspecting data

### Non-tidy way

To preview the data to see what it looks like briefly, `head()` is a common function that returns the first n rows/elements of a vector, dataframe, etc.

```{r}
head(wm_raw_untidy)
```

What happens if you forget to call `head()` and just print the whole dataframe?

```{r}
wm_raw_untidy
```

As Charles Barkley would say, turrible. Do you really need to look at the whole dataframe like this?

You may also step up one from `head()` and call `str()`:

```{r}
str(wm_raw_untidy)
```

Also, take a moment to talk about the Viewer--I myself used to be a stan for the Viewer but the key issue with the Viewer is that it becomes slow to load for large objects and is unwieldy for visual comparison of data in objects with very many rows and/or columns.

### Tidy way

`tibble` is LIFE

```{r}
raw
```

How does she print this way? She is a tibble and she is beautiful. Note that packages within the tidyverse work together--raw rectangular data read in using `readr` is given the tibble class by default.

```{r}
class(raw)
```

`"tbl_df"` is the class name for tibbles.

```{r}
class(wm_raw_untidy)
```

Dataframes can be coerced to tibble, though, if they have been created using base R.

```{r}
# This will give us the sweet sweet print method. It's like str() but better. And happens by default!
as_tibble(wm_raw_untidy)
```

### Key differences

Compared to a dataframe, a tibble:

* has a cleaner & more informative `print()` method
* is stricter when you create tibbles using `tibble()`, or when you index (with [] or $)

There are a couple other useful things that tibbles do that dataframes don't, but the strictness and the print method are the two most useful things that I find. As you'll see, imposing a bit more strictness on R can have the beneficial side effect of minimizing buggy cases in your own code, since you're forced to be more explicit.

## Subsetting data

We'll deal with subsetting dataframes two ways: choosing a subset of columns and a subset of rows.

### Tidy way

#### Subsetting by column

Note that these are **unquoted** variable names and not a character vector.

Also, just some nice shit that you can do with `select()` etc

```{r}
# This auto-indenting shit after the pipe works because the pipe is a left-right operator
# Renaming columns is so smooth in the tidyverse!
raw %>%
  select(id, age, gender, trial_num, is_old, response, rt)
```

```{r}
# Renaming columns is so smooth in the tidyverse!
raw %>%
  select(subject_id = id, age, gender, trial_num, is_old, response, rt)
```

```{r}
raw %>%
  select(id, starts_with("q"))
```


#### Subsetting by row

Notice that we no longer have to index the original dataframe when specifying the logical vector(s) that will be used to subset rows

```{r}
raw %>%
  filter(id == 13)
```

Another function that technically "subsets" data, but not logically: `distinct()` is the `dplyr` function for removing rows with duplicate observations

```{r}
raw %>%
  select(id, gender, age) %>%
  # the .keep_all argument specifies whether to drop columns that aren't being distinct-ed by
  distinct(id, .keep_all = TRUE)
```


#### Subsetting by row and column simultaneously

In this case, order doesn't super matter, so it's a matter of personal preference. I would generally recommend putting all renaming calls as early as possible to allow the largest block of code to be consistent with itself.

```{r}
raw %>%
  select(id, age, gender, trial_num, is_old, response, rt) %>%
  filter(id == 13)
```

```{r}
# dplyr::filter() takes logical statements with & and | in them like you might use in bracket indexing,
# but it also has a nice feature: specifying separate logical statements as separate args,
# separate by commas, implicitly acts like chaining the statements together with &
# since the intersection of logical vectors is probably the most common use case for row subsetting
raw %>%
  select(id, age, gender, trial_num, is_old, response, rt) %>%
  filter(id == 13, is_old == 1)
```

### Key differences

Bracket indexing always happens from the global environment, while `dplyr` column/row subsetting verbs all act first within the local environment of the tibble of interest. For example, this means that when specifying columns for logical subsetting, columns can be named bare, without referencing the home tibble using $. This minimizes typing and also risk of accidentally referencing objects you didn't mean to reference.

## Reshaping data 

### Non-tidy way

Honestly, there are so many ways people can do this using older code--`reshape`, `reshape2`, etc etc.

```{r}

```


### Tidy way

`tidyr` goes here, maybe. Since the example starts with trialwise data, maybe not yet? Should wait until we get to summary statistics?

## Manipulating data

`dplyr` goes here

### Non-tidy way

### Tidy way

```{r}
wm_tidy <- wm_raw_tidy %>%
  select(subjNum, groupStatus, numDistr = NumDistr, cond = Cond, resp, rt = RT) %>%
  # A tour of useful functions of sorts
  # For turning anything into a factor variable, special case of a switch statement where every case in the input has only one case in the output
  mutate(cond = recode_factor(cond,
                              `0` = "no_change",
                              `1` = "change"),
         # case_when: more general switch statement, can take any logical vector of appropriate length as a condition
         # remember that all tidyverse functions are evaluated sequentially, so these logical statements are hierarchical per an if-else if situation
         resp = case_when(resp == "j" ~ "no_change",
                          resp == "jj" ~ "no_change",
                          resp == "k" ~ "change",
                          resp == "kk" ~ "no_change",
                          TRUE ~ NA_character_),
         # nice function from the forcats pkg to reorder factor levels
         resp = fct_relevel(resp, c("no_change", "change")),
         # slightly more complex logical calls in case_when
         acc = case_when(cond == "change" & resp == "change" ~ "hit",
                         cond == "change" & resp == "no_change" ~ "miss",
                         cond == "no_change" & resp == "no_change" ~ "cr",
                         cond == "no_change" & resp == "change" ~ "fa",
                         TRUE ~ NA_character_),
         # coalesce is nice to fill all NAs in a vector with some other value
         acc_no_na = coalesce(acc, "no_response"),
         # conversely, should you need to replace a certain value in a vector with NA
         rt = na_if(rt, 0)) %>%
  # enter tidyverse GROUPING, which causes vectorized functions that would ordinarily operate over the whole length of the tibble to operate independently over GROUPS of the tibble defined by all visible levels of grouping variable
  group_by(groupStatus, subjNum, numDistr) %>%
  # one thing you can do with grouped tibbles is feed them directly into nest() and the grouping columns will become the label columns in the nested tibble
  # list-columns-of-tibbles!!!
  nest(.key = "data") %>%
  # this is the tibble-sorting function
  arrange(subjNum, numDistr) %>%
  # map()!!! we'll get into this
  # esp because summarize() is nested inside here
  mutate(overall = map(data, ~.x %>%
                         summarize(rate_hit = sum(acc_no_na == "hit") / sum(cond == "change"),
                                   rate_fa = sum(acc_no_na == "fa") / sum(cond == "no_change")) %>%
                         mutate(k = 2 * (rate_hit - rate_fa)))) %>%
  # to get out of list-column-of-tibbles format and back into one-long-tibble 
  unnest(overall, .preserve = "data")
```

```{r}
ltm_tidy <- ltm_raw_tidy %>%
  select(subjNum, groupStatus, numDistr = NumDistractorsStudy, oldNew = OldNew, resp = recogresp, rt = recogRT) %>%
  mutate(oldNew = recode_factor(oldNew,
                                New = "new",
                                Old = "old"),
         resp = case_when(resp == "j" ~ "old_high",
                          resp == "jj" ~ "old_high",
                          resp == "k" ~ "old_low",
                          resp == "kk" ~ "old_low",
                          resp == "f" ~ "new_high",
                          resp == "ff" ~ "new_high",
                          resp == "d" ~ "new_low",
                          resp == "dd" ~ "new_low",
                          TRUE ~ NA_character_)) %>%
  separate(resp, into = c("resp", "confidence")) %>%
  mutate(acc = case_when(oldNew == "old" & resp == "old" ~ "hit",
                         oldNew == "old" & resp == "new" ~ "miss",
                         oldNew == "new" & resp == "new" ~ "cr",
                         oldNew == "new" & resp == "old" ~ "fa",
                         TRUE ~ NA_character_),
         acc_no_na = coalesce(acc, "no_response"),
         rt = na_if(rt, 0)) %>%
  group_by(groupStatus, subjNum) %>%
  nest(.key = "data") %>%
  arrange(subjNum) %>%
  mutate(overall = map(data, ~.x %>%
                         summarize(rate_hit = sum(acc_no_na == "hit") / sum(oldNew == "old"),
                                   rate_fa = sum(acc_no_na == "fa") / sum(oldNew == "new")))) %>%
  unnest(overall, .preserve = "data")
# given that false alarms are all marked with numDistr == 0 may want to store this a slightly different way
```
